OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-06-28 paper/paper.Rmd
Describe MCC calculation

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiClust first seeds each sequence into its own OTU as a singleton. 
Then for each sequence, OptiClust considers whether the sequence should move to
a different OTU or remain in its current OTU, choosing the option that results
in a better Matthews correlation coefficient (MCC).
**TODO: describe how it generates a confusion matrix and calculates the MCC.**
Iterations continue until the MCC stabilizes or until a maximum number of
iterations is reached.
This process produces _de novo_ OTU assignments with the most optimal MCC given
the input sequences.
OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03).
Initially, all query sequences are placed into separate OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the MCC. 
Alternatively, a sequence will remain unassigned if the MCC value is maximized
when the sequence is a singleton rather than assigned to a reference OTU.
This process is repeated until the MCC changes by no more than 0.0001 (default)
or until a maximum number of iterations is reached (default: 100).
In the closed reference mode, any query sequences that cannot be assigned to
references OTUs are discarded, and the results will only contain OTUs that exist
in the original reference.
In the open reference mode, unassigned query sequences are clustered _de novo_
using OptiClust to generate new OTUs.
The final MCC is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 
1) fit the query sequences to reference OTUs generated by _de novo_ clustering an
independent database, or
2) split the dataset into a reference and query fraction, cluster the reference
sequences _de novo_, then fit the query sequences to the reference OTUs.
We clustered sequences from four datasets isolated from soil
[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse gut
[@schloss_stabilization_2012], and human gut [@baxter_microbiota-based_2016]
samples to test the performance of OptiFit with both of these strategies.

>>>>

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiClust first seeds each sequence into its own OTU as a singleton. 
Then for each sequence, OptiClust considers whether the sequence should move to
a different OTU or remain in its current OTU, choosing the option that results
in a better Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
The MCC uses all values from a confusion matrix and ranges from zero to one,
with a score of one occurring when all sequence pairs are true positives and
true negatives, and a score of zero when all pairs are false positives and false
negatives.
Sequence pairs that are similar to each other (i.e. within the distance
threshold) are counted as true positives if they are assigned to the same OTU,
and false negatives if they are not in the the same OTU.
Sequence pairs that are not similar to each other are true negatives if they are
not assigned to the same OTU, and false positives if they are not in the same
OTU.
OptiClust iterations continue until the MCC stabilizes or until a maximum number
of iterations is reached.
This process produces _de novo_ OTU assignments with the most optimal MCC given
the input sequences.

OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03).
Initially, all query sequences are placed into separate OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the MCC. 
Alternatively, a sequence will remain unassigned if the MCC value is maximized
when the sequence is a singleton rather than assigned to a reference OTU.
All query and reference sequence pairs are considered when calculating the MCC.
This process is repeated until the MCC changes by no more than 0.0001 (default)
or until a maximum number of iterations is reached (default: 100).
In the closed reference mode, any query sequences that cannot be assigned to
references OTUs are discarded, and the results will only contain OTUs that exist
in the original reference.
In the open reference mode, unassigned query sequences are clustered _de novo_
using OptiClust to generate new OTUs.
The final MCC is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 
1) fit the query sequences to reference OTUs generated by _de novo_ clustering an
independent database, or
2) split the dataset into a reference and query fraction, cluster the reference
sequences _de novo_, then fit the query sequences to the reference OTUs.
We clustered sequences from four datasets isolated from soil
[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse gut
[@schloss_stabilization_2012], and human gut [@baxter_microbiota-based_2016]
samples to test the performance of OptiFit with both of these strategies.
